---
title: "Stack Overflow Question - Znew_gen2"
author: "Hillary Heiling"
date: "February 18, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# How to fix a memory allocation error when calling a C++ function in a compiled R package?

I have a C++ function that works when sourced independently, but when I include it in a compiled R package, I keep receiving the following error: `error: arma::memory::acquire(): out of memory`. In this post, I have provided the C++ code containing the function `Znew_gen2` that I want to use in my compiled R package. Using a working example, I can show that the `Znew_gen2` function works when I source it independently (outside of an R package). However, when I call this same function from an example R package named `HelpWithZnewgen2`, the same example no longer works and I get the error `error: arma::memory::acquire(): out of memory`. All of the code mentioned in this post can be found in the GitHub repo <https://github.com/hheiling/myrepo_R>.

My C++ function, named `Znew_gen2`, is provided below:

```{r, echo=FALSE}
#define ARMA_64BIT_WORD
#include <RcppArmadillo.h>
// [[Rcpp::depends(BH, bigmemory, RcppArmadillo)]]
#include <bigmemory/MatrixAccessor.hpp>
#include <numeric>

using namespace Rcpp ;

// [[Rcpp::export]]
void Znew_gen2( const arma::mat& U, const arma::mat& Z, const arma::vec& g, const arma::vec& cols, 
                unsigned int n, unsigned int q, unsigned int d, SEXP pBigMat, 
                arma::sp_mat& J){ 
  
  unsigned int i = 0;
  unsigned int j = 0;
  unsigned int nMC = U.n_rows;
  unsigned int index = 0;
  int gr = 0;
  arma::mat Usub(1, q);
  arma::mat Zsub(1, q);
  arma::mat out(n*nMC, J.n_cols);
  XPtr<BigMatrix> pMat(pBigMat); 
  MatrixAccessor<double> out2(*pMat); 
  
  Rprintf("First For Loop \n");
  for(i = 0; i<n; i++){
    gr = g(i);
    for(index = 0;index<q; index++){
      Zsub(0,index) = Z(i,cols(index)-1 + gr - 1);
    }
    for(j = 0; j < nMC; j++){
      for(index = 0;index<q; index++){
        Usub(0,index) = U(j,cols(index)-1 + gr - 1);
      }
      out.row((i*nMC + j)) = kron(Usub, Zsub) * J;
    }
  }
  
  Rprintf("Second For Loop \n");
  for(i = 0; i<n*nMC; i++){
    for(j = 0; j<out.n_cols; j++){
      out2[j][i] = out(i, j);
    }
  }
  
  Rprintf("End of Function \n");
}

```

This function can be found in the `Znew_gen2.cpp` file in the GitHub repo.

This file can be sourced as follows:

```{r}
library(Rcpp)

# Code to download the "Znew_gen2.cpp" file from the GitHub repo  
## Specify where you want the file to download to:
destination_file = "Source_Code_Znew_gen2.cpp" 
    # Can rename if you like, but must not have spaces in the filename
download.file(url = "https://raw.githubusercontent.com/hheiling/myrepo_R/master/Znew_gen2.cpp", 
              destfile = destination_file)
sourceCpp(file = destination_file)

```

My working example is provided here:

```{r}
# Working Example:

library(bigmemory)
library(Matrix)

set.seed(1)
U = matrix(rnorm(3000), nrow=100, ncol=30)
Z = matrix(rnorm(15000), nrow = 500, ncol = 30)
group = rep(1:10, each = 50)
cols = c(1,11,21)
n = 500
q = 3
d = 10 
Znew = big.matrix(nrow = nrow(Z)*nrow(U), ncol = 6)
J_SpMat = Matrix(0, 9, 6, sparse = TRUE)
  sumy = 0
  sumx = 0
  zeros = 0
  for(i in 1:3){
    J_SpMat[ sumx + zeros + 1:(3 - (i-1)), sumy + 1:(3 - (i-1))] = diag((3 - (i-1)))
    sumy = sumy + (3 - (i-1))
    sumx = sumy
    zeros = zeros + i
  }

```

If I run the code given above and then call the sourced `Znew_gen2` function, the `Znew_gen2` function runs and the lines "First For Loop", "Second For Loop", and "End of Function" appear, which indicates that the function ran through all of the lines of the `Znew_gen2` code without error.

```{r}
# Calling the sourced `Znew_gen2` function:
Znew_gen2(U, Z, group, cols, n, q, d, Znew@address, J_SpMat)
```

I now want to include this function in an R package, but as I have it currently, the function doesn't work when called in this setting. For illustrative purposes, I have created an R package called `HelpWithZnewgen2`, which is provided on the GitHub repo. This R package can be downloaded using the following code:

```{r}
library(devtools)
library(remotes)
install_github("hheiling/myrepo_R", subdir = "HelpWithZnewgen2")

library(HelpWithZnewgen2)
```

In order to call this `Znew_gen2` function within the R package, I use the following wrapper function (found in the file `HelpWithZnewgen/R/wrapper_functions.R` within the R package):

```{r}
#' @useDynLib HelpWithZnewgen2
#' @export
Znew_gen2.Rfunction = function(U, Z, group, cols, n, q, d, pBigMat, J){
  Znew_gen2(U, Z, group, cols, n, q, d, pBigMat, J)
}
```

Now when I call the `Znew_gen2.Rfunction` function from the complied R package (using the same working example provided previously), I get the following error:

```{r}
# Calling the function from the compiled package:
Znew_gen2.Rfunction(U, Z, group, cols, n, q, d, Znew@address, J_SpMat)
```

```{r}
## Error received:
# error: arma::memory::acquire(): out of memory
# Error in Znew_gen2(U, Z, group, cols, n, q, d, pBigMat, J) : 
#   std::bad_alloc
```

Because the code itself works when sourced outside of the R package, I suspect my issue is related to how my R package is set up. From searching online, I wouldn't be suprised if there is an issue with one or more of the following components: the lines in `Znew_gen2.cpp` before the "using namespace Rcpp" line, my Description file, or possibly some line I am missing in the `Makevars` file of the R package. Despite my suspicions (which may not be correct, since I'm relatively new to writing R packages), I haven't been able to address the issue. Therefore, I would greatly appreciate any suggestions on how to fix this problem. 

I have included the following lines in my `Description` file:

```{r, eval=FALSE}
Package: HelpWithZnewgen2
Type: Package
Title: Need Advice for Compiling and Running Znew_gen2 Function Properly
Version: 1.0
Date: 2019-02-01
Authors@R: c(
    person("Hillary", "Heiling", email = "hheiling@live.unc.edu", role = c("aut", "cre")),
    person("Naim", "Rashid", email = "naim@unc.edu", role = c("aut")))
Author: Who wrote it
Maintainer: Who to complain to <yourfault@somewhere.net>
Description: More about what it does (maybe more than one line)
License: What license is it under?
Depends: lme4, irr, ncvreg, mvtnorm, ROCR, Rcpp,
        RcppArmadillo, inline, grpreg, bigmemory, bigalgebra, BH,
        glmnet
LinkingTo: Rcpp, RcppArmadillo, bigmemory, BH
RoxygenNote: 6.1.0.9000
```

